Collaborative Learning with Dirichlet Process Clustering for Rapid Online Adaptation in Robotics

### 2023-09-06 paper/methodology.tex
Sep 6

\section{Methodology}
We are aiming to improve RTE by leveraging the collaboratively collected real-world interactions.
To be clear about the terminology, in repertoire-based method, an interaction means executing an certain elementary policy and acquiring its outcome.
DP enables us to achieve this by extending the GP model in RTE to an infinite mixture of GPs trained with real-world data.
Previous works of training infinite mixture of GP via DP have been made in \cite{infinite_MGP, variational_MGP}. 
But in our case, the design needs to be different.
First, our real-world interaction data are produced in individual episodes, where each episode corresponds to the interactions collected during each deployment.
Since we know that each deployment is made in a certain environment, all the interactions in an episode should correspond to the same dynamics. 
Hence, when clustering the historical data we need to conduct clustering on the episode but not on the individual data. 
The second difference from previous works is that we are decoupling each dimension and each episode.
The decoupling of dimension is inherited from RTE, while the decoupling of episodes is a unique design.
This is based on the fact that data within each episode come from the same environment and are hence correlated, while different episodes do not necessarily correspond to the same environment even they are grouped into the same cluster.
Such design allows us to group up environments that are different but similar, providing higher flexibility.
Another reason for using decoupled episodes is to make our system scalable.
Since we are now clustering episodes which typically contains dozens of data, each clusters can have thousands of data.
GP requires to compute the inverse of the joint covariance matrix, which has time complexity scaling cubic with data size.
Such computational expense is completely unaffordable for online adaptation. 
If the episodes are decoupled, the joint covariance matrix becomes a block diagonal matrix, and the time complexity of its inverse scales only linearly.

To really implement the DP clustering, there are many remaining challenges to be addressed.
The prior mean and kernel are the two most important factors of a GP.
It can be very hard to design a good prior function or to determine the length scales as we have little knowledge about the real-world dynamics.
A few parts of our methodology heavily rely on simulation.
For this, the simulation has to be relatively accurate and free of bugs.
We also make a very strong assumption that the simulated results always correspond to real-world outcomes in some environment. 
For example, we set the friction coefficient to 0.9 in the simulation, but the friction is underestimated and the simulated outcomes are very close to the real-world results for friction coefficient equal to 0.8.
Hence we believe that although the simulated results are not strictly accurate, they can still represent some real-world results.
This assumption not always true, but it plays a very important role in the methodology.

@@

In GP, the prior mean function gives a prior estimate of the outcomes of all the elementary policies in a certain environment.
To study the dynamics, we express its dependency in the following form:

@@

Combined with the assumption we made about the simulation, Eq. (\ref{linear_form}) suggests we may approximate the real-world distortion of the entire repertoire using a linear model.

@@

\subsection{Clustering the Dynamics with Dirichlet Processes}
So far, we have determined the basis for our linear prior mean function, the input space, different representations of dynamics as well as the way to calculate relative noises.
To cluster the historical data with DP, we still need the prior distribution $H$ for the detailed parameters. 
In our case where the dynamics are modelled by GP, these parameters include the weights $\bm{w}$ for the linear mean function, the length scales $\bm{l}$ and the prior variance $\alpha_0$ for the kernel, and the noises $\sigma_{n}^2$ for covariance matrix.

This is hard to do since we have little knowledge.
Even if we could give a perfect prior distribution, the integrals in Eq. (\ref{integral_1}) and Eq. (\ref{integral_2}) are intractable.
The solution is use Monte Carlo sampling in simulation combined with a trick called posterior convergence estimate.
We can see that these parameters are all related to the environment.
The Monte Carlo sampling is to collect a sample of these environment-related parameters from the simulation as our prior distribution $H$.
To do this, we first randomize the simulation settings to get a sample of simulated environments.
We then collect a sufficient amount (e.g. 64) of interaction data from each environment and find the detailed parameters using maximum likelihood estimate (MLE):
\begin{equation}
\bm{\theta}^* = 
\arg \max_{\bm{\theta}} \, p(\bm{y}|\bm{\theta})
= \arg \min_{\bm{\theta}} \, -\log p(\bm{y}|\bm{\theta})
\label{MLE}
\end{equation}
The Monte Carlo sampling seems computationally expensive. But note that the sample amount for each environment is small and invariant to the size of the repertoire.
Hence it is much cheaper to do than finding the basis for the prior mean function, where we need to evaluate all the policies in several environments.

The probabilistic model in our case is a product of GPs over the dimensions, and $p(\bm{y}|\bm{\theta})$ can be calculated from Eq. (\ref{multi_normal}):
\begin{equation}
\begin{gathered}
-\log p(\bm{y}|\bm{\theta}) = \sum_{d} -\log p(\bm{x}_{d}|\bm{\theta}_d)
\\
= \frac{1}{2} \sum_{d}
(\bm{x}_{ d} - \bm{\mu}_{\bm{\theta}_d})^T \bm{\Sigma^{-1}}_{\bm{\theta}_d} (\bm{x}_{d} - \bm{\mu}_{\bm{\theta}_d})
\\ + \sum_{ d} \left[ \frac{N}{2}\log(2\pi) 
+ \log \det(\bm{\Sigma}_{ \bm{\theta}_d}) \right]
\end{gathered}
\label{nll}
\end{equation}
where the weights $\bm{w}_d$ are encoded in $\bm{\mu}_{\bm{\theta}_d}$, and the rest of the parameters are encoded in $\bm{\Sigma}_{\bm{\theta}_d}$.
It is worth noting that the noise $\sigma_n^2$ in displacement-based representation is the same for all policies; while in arc-based representation, Eq. (\ref{covariance_matrix}) needs to be modified:
\begin{equation}
\bm{\Sigma} = \bm{K}(\bm{x_{1:N}}, \bm{x_{1:N}}) + \sigma_{n}^2 \bm{D}
\label{modified_covariance_matrix}
\end{equation}
where matrix $\bm{D}$ is a diagonalized matrix with entries calculated according to Eq. (\ref{noises_for_v_w_phi}).
The optimal solution of Eq. (\ref{MLE}) can be obtained using optimization algorithms like L-BFGS-B. The integrals in Eq. (\ref{integral_2}) and Eq. (\ref{integral_1}) hence become the averages of finite number of terms:
\begin{equation}
\begin{gathered}
p(y_i|c_i, \bm{y}_{-i}, \bm{c}_{-i}) = 
\sum_{l=1}^L p(y_i|\bm{\theta}^{(l)})
p(\bm{\theta}^{(l)}|c_i, \bm{y}_{-i}, \bm{c}_{-i})
\\
p(y_i|c_i \neq j \, \text{for any} \, n_j \neq 0) = 
\frac{1}{L} \sum_{l=1}^L p(y_i|\bm{\theta}^{(l)})
\end{gathered}
\label{MC_prior_integral}
\end{equation}
Where the posterior distribution is:
\begin{equation}
p(\bm{\theta}^{(j)}|c_i, \bm{y}_{-i}, \bm{c}_{-i}) = 
\frac{
\prod_{c_k=c_i, k \neq i} p(y_k|\bm{\theta}^{(j)})}
{\sum_{l=1}^L \prod_{c_k=c_i, k \neq i} p(y_k|\bm{\theta}^{(l)})}
\label{MC_posterior}
\end{equation}
In practice, it is much easier and numerically stable to work with logarithm likelihoods as in Eq. (\ref{nll}), as the direct product of likelihoods will very likely lead to overflow.

Although the integrals are now made tractable, this method has two fatal shortages.
First, the Monte Carlo sampling only provides limited precision of the parameters.
As each cluster grow larger, Eq. (\ref{MC_posterior}) implies that the set of parameters that is closest to the ground truth will eventually make up 100\% of the posterior weight.
Hence, the number of clusters will not exceed the amount of sampled priors. 
Second and more importantly, the prior collection is made in simulation, which ensures no extrapolation to the real-world.
To solve these two shortages, we introduce a trick called posterior convergence estimate.
As a cluster gathers more data, the posterior distribution of its parameters will become narrower and more centred closer to the ground truth (see Fig \ref{posterior_convergence}).
\begin{figure}[h]
\centering
\includegraphics[width=0.45\textwidth]{posterior_convergence.pdf}
\caption{Example of the convergence of parameter posterior in DPMM.
Data are sampled from a infinite mixture of Gaussians, where the variance is fixed at 4, and the mean is given by a Gaussian mixture as the prior.
As the cluster grows larger, the posterior of the mean becomes more narrowed and centred at the ground truth (5.5).
}
\label{posterior_convergence}
\end{figure}
%
When the cluster size is large, we can safely assume that the posterior becomes a sharp spike.
Hence, we may treat it as a Dirac delta distribution, and Eq. (\ref{integral_1}) becomes:
\begin{equation}
\begin{gathered}
p(y_i|c_i, \bm{y}_{-i}, \bm{c}_{-i}) \approx 
\int p(y_i|\bm{\theta})
\delta (\bm{\theta} - \bm{\theta}^*)
d\bm{\theta}
\\
= p(y_i|\bm{\theta}^*)
\end{gathered}
\label{posterior_convergence_estimate}
\end{equation}
This is called posterior convergence estimate, and the center of this Dirac delta is located using MAP. 
This means we refit the parameters for large clusters (with data size above a threshold, e.g. 30) as their posterior.
This design is to leverage the fact that for small clusters the posteriors haven't converged, and their integration can be efficiently approximated by Eq. (\ref{MC_prior_integral}).
While for large clusters, their posterior cannot be efficiently represented by any prior, so we allow them to determine the best parameters based on their data.
The only thing that such prior should satisfy is that it covers a variety of possible cases.
Thus, the use of Monte Carlo sampling for the prior acts like a guide for smaller clusters, and neither the coverage of parameters or the number of clusters is constrained by the prior.

Because we have no access to the analytical form of the prior to do MAP, we can use MLE instead (for large data size, MLE and MAP gives similar results).
Hence, during Gibbs sampling, the likelihood of an episode belonging to a large cluster is calculated according to Eq. (\ref{posterior_convergence_estimate}), while the likelihood of belonging to a small cluster is calculated using the first equation in Eq. (\ref{MC_prior_integral}).
As mentioned previously, the episode needs to be removed from its current cluster which will affect the posterior when calculating the likelihood of this episode staying in its current cluster.
For small cluster, we can recalculate the posterior using Eq. (\ref{MC_posterior}).
For large cluster, this means the parameter $\bm{\theta}^*$ needs to refitted with this episode removed, if the data size is still large enough without this episode. 
After each resample step, we also need to refit the parameters for large clusters if any change is made.
Be aware that the subscript $i$ in Eq. (\ref{MC_prior_integral}), Eq. (\ref{MC_posterior}) and Eq. (\ref{posterior_convergence_estimate}) refers to each episode instead of each individual interaction data. 
Similarly, the $n$ in Eq. (\ref{indicator_posterior_3}) corresponds to the number of episodes instead of the amount of data.

In practice, refitting all the parameters in each step of Gibbs sampling could lead to performance issues as the Gibbs sampling cannot be paralleled.
It is much cheaper to just refit the weights for the prior mean function and leave the rest of the parameters fixed.
It is reasonable as the prior is most important and can be obtained analytically:
\begin{equation}
\bm{w}_d^* = \left(\sum_{i=1}^n
\bm{X}_i^T \bm{\Sigma}_{i, \bm{\theta}_d}^{-1} \bm{X}_i
\right)^{-1} \sum_{i=1}^n \bm{X}^T_i \bm{\Sigma}_{i, \bm{\theta}_d}^{-1} \Delta \bm{x}_{i, d}
\label{refitted_weights}
\end{equation}
where the summation is made over all the episodes in the cluster.
$\Delta \bm{x}_{i, d}$ is the real-world distortions of the $i^{\text{th}}$ episode.
$\bm{X}_i$ is the coordinate matrix of this episode, where each row gives the coordinates of the policy corresponding to each interaction.
Since the rest of the parameters are fixed, the covariance matrix $\bm{\Sigma}_{i, \bm{\theta}_d}$ is generated using the most likely prior sample.

\subsection{Learning the Dynamics Online}
After having clustered the historical data, we can use the result to assist adaptation.
In RTE, we learn the distortion online using GPs, where the parameters (prior mean function, kernel etc.) are determined manually.
In our method, the dynamics is modelled with an infinite mixture of GPs. 
The GP parameters are given by the posterior distribution found with historical data, which is a mixture model of finite terms:
\begin{equation}
p(\bm{\theta}) = \sum_j \frac{n_j}{n + \alpha} p(\bm{\theta}|c=j) 
+ \frac{\alpha}{n + \alpha} H(\bm{\theta})
\label{trained_mixture_model}
\end{equation}
where the summation is made over the clusters containing at least one episode, and $p(\bm{\theta}|c=j)$ denotes the posterior for the $j^{\text{th}}$ cluster.
Each instance of $\bm{\theta}$ sampled from $p(\bm{\theta})$ is a certain configuration of parameters of GP.
The last term results from the important property of DPMM that there is always a non-zero probability that we will get a new cluster (face a new situation).
The online learned distortion is integrated over the posterior of $\bm{\theta}$ conditioned on the collected real-world interactions:
\begin{equation}
\begin{gathered}
p(x^*_d|\bm{x}_{1:N}) 
\\
= \sum_j p(c=j| \bm{x}_{1:N}) \int p(x^*_d|\bm{\theta}, \bm{x}_{1:N})p(\bm{\theta}|c=j, \bm{x}_{1:N}) d\bm{\theta} 
\\
+ p(c \neq j, \forall j|\bm{x}_{1:N}) \int p(x^*_d|\bm{\theta}, \bm{x}_{1:N}) p(\bm{\theta} | \bm{x}_{1:N}) d\bm{\theta}
\end{gathered}
\label{MGP_posterior}
\end{equation}
where the indicator probability is given by:
\begin{equation}
\begin{gathered}
p(c=j| \bm{x}_{1:N}) 
= \frac{n_j \cdot p(\bm{x}_{1:N}|c=j)
}{Z}
\\ 
p(c \neq j, \forall j| \bm{x}_{1:N})
= \frac{\alpha \cdot p(\bm{x}_{1:N}|c \neq j, \forall j)
}{Z}
\end{gathered}
\label{MGP_indicator}
\end{equation}
where $\bm{x}_{1:N}$ stands for the collected ($N$ number of) interaction data during deployment.
This is complicated and requires calculating multiple integrals over the posteriors. 
Fortunately, if we have applied posterior convergence estimate for $p(\bm{\theta}|c=j)$, we will have:
\begin{equation}
\begin{gathered}
\int p(x^*_d|\bm{\theta}, \bm{x}_{1:N})p(\bm{\theta}|c=j, \bm{x}_{1:N}) d\bm{\theta} \approx p(x^*_d|\bm{\theta}^*_j, \bm{x}_{1:N})
\\
\text{and } p(\bm{x}_{1:N}|c=j) \approx p(\bm{x}_{1:N}|\bm{\theta}^*_j)
\end{gathered}
\label{posterior_convergence_simplification}
\end{equation}
where $\bm{\theta}^*_j$ denotes the optimal parameters for the cluster $j$.
In actual implementation, the clustering of historical data with DP is conducted in the archive management system where expensive data processing is available. 
The results are saved as an adaptation system to be used (downloaded) by individual robots.
The adaptation system is independent from the archive, and we are free to make simplifications for embedded systems.
Hence, we can apply posterior convergence estimate for all the clusters we have found for simplicity.
For clusters that are not large enough to refit the parameters, we will just use the parameters of the most likely prior sample.
This might be inappropriate for clusters that are very small, and hence we can discard them from the adaptation system.

>>>>

\section{Methodology}
We are aiming to improve RTE by leveraging the collaboratively collected real-world interactions as the priors of our GP.
The greatest difficulty lies in the way to build the priors.
In GPCF, data from each previous user is treated as one of the candidate priors.
If we are to use the same design, we can use the posteriors calculated (with GP) using the data collected from each single task as the priors.
However, this would not be a very ideal design in our case.
The first reason is that we will be having a large number of priors after we have collected many data.
Second, it is likely that some data many come from similar dynamics and should be grouped together to enable more sophisticated analysis.
Hence, we took a different approach by first clustering the data based on their dynamics and then building a prior for each cluster.
Since the dynamics for each previous deployment is modelled by GP, we can use DP to group them in a non-parametric manner.
The result of such DP clustering will be an infinite mixture of GPs, where each cluster will be a GP with the prior and parameters following a certain posterior distribution.
We then replace the transformation model in RTE by this infinite mixture of GPs, and let the robot identify the most likely GP during real-world adaptation.

Previous works of training infinite mixture of GP using DP clustering have been made in \cite{infinite_MGP, variational_MGP}. 
But in our case, the design needs to be different.
First, our real-world interaction data are produced in individual episodes, where each episode corresponds to the interactions collected during each deployment.
Since we know that each deployment is made in a certain environment, all the interactions in an episode should correspond to the same dynamics. 
Hence, when clustering the historical data we need to conduct clustering on the episode but not on the individual data. 
The second difference from previous works is that we are decoupling each dimension and each episode.
The decoupling of dimension is inherited from RTE, while the decoupling of episodes is a unique design.
This is based on the fact that data within each episode come from the same environment and are hence correlated, while different episodes do not necessarily correspond to the same environment even they are grouped into the same cluster.
Such design allows us to group up environments that are different but similar, providing higher flexibility.
Another reason for using decoupled episodes is to make our system scalable.
Since we are now clustering episodes which typically contains dozens of data, each clusters can have thousands of data.
GP requires to compute the inverse of the joint covariance matrix, which has time complexity scaling cubic with data size.
Such computational expense is completely unaffordable for online adaptation. 
If the episodes are decoupled, the joint covariance matrix becomes a block diagonal matrix, and the time complexity of its inverse scales only linearly.

It can be very hard to design a good prior function or to determine the kernel if we have completely no idea about the real-world dynamics.
Hence, a few parts of our methodology relies on the use of simulation.
For this, the simulation has to be relatively accurate and free of bugs.
We also make a very strong assumption that the simulated results always correspond to real-world outcomes in some environment. 
For example, we set the friction coefficient to 0.9 in the simulation, but the friction is underestimated and the simulated outcomes are very close to the real-world results for friction coefficient equal to 0.8.
Hence we believe that although the simulated results are not strictly accurate, they can still represent some real-world results.
This assumption not always true, but it plays a very important role in the methodology.

@@

In GP, the prior mean function is most important as it provides the prior estimate of the outcomes of all the elementary policies.
To determine a suitable form of the prior mean function, we hope to study the dependencies of the dynamics.
We know that the outcome of a policy should depend on the behaviour of this policy as well as the environment.
Hence, we express the dynamics in the following form:

@@

Combined with the assumption we made about the simulation, Eq. (\ref{linear_form}) suggests we may approximate the real-world distortion of the entire repertoire using a linear model, which is consistent with the design in GPCF.

@@

\subsection{Clustering the Dynamics with Dirichlet Processes}
So far, we have determined the basis for our linear prior mean function, the input space, different representations of dynamics as well as the way to calculate relative noises.
To cluster the historical data with DP, we still need the prior distribution $H$ for the detailed parameters for the GPs. 
These parameters typically refer to the weights $\bm{w}$ for the linear mean function.
But in our case, we also allowed the kernels to be non-stationary for higher flexibility.
This design is to incorporate the possibility that the Bayesian inference in the input space could be different in each environment.
Hence, we also need to determine the prior distribution for the length scales $\bm{l}$, the prior variance $\alpha_0$, and the noises variance $\sigma_{n}^2$.
Such distribution is hard to give since we have little prior knowledge.
Even if we could give a perfect prior distribution, the integrals in Eq. (\ref{integral_1}) and Eq. (\ref{integral_2}) are intractable since we allowed the kernel parameters to be variables.

The solution is to use Monte Carlo sampling combined with a trick called posterior convergence estimate.
We can see that these parameters are all related to the environment.
The Monte Carlo sampling is to collect a sample of these environment-related parameters from the simulation as our prior distribution $H$.
To do this, we first randomize the simulation settings to get a sample of simulated environments.
We then collect a sufficient amount (e.g. 64) of interaction data from each environment and find the detailed parameters using maximum likelihood estimate (MLE):
\begin{equation}
\bm{\theta}^* = 
\arg \max_{\bm{\theta}} \, p(\bm{y}|\bm{\theta})
= \arg \min_{\bm{\theta}} \, -\log p(\bm{y}|\bm{\theta})
\label{MLE}
\end{equation}
The probabilistic model in our case is a product of GPs over the dimensions, and $p(\bm{y}|\bm{\theta})$ can be calculated from Eq. (\ref{multi_normal}):
\begin{equation}
\begin{gathered}
-\log p(\bm{y}|\bm{\theta}) = \sum_{d} -\log p(\bm{x}_{d}|\bm{\theta}_d)
\\
= \frac{1}{2} \sum_{d}
(\bm{x}_{ d} - \bm{\mu}_{\bm{\theta}_d})^T \bm{\Sigma^{-1}}_{\bm{\theta}_d} (\bm{x}_{d} - \bm{\mu}_{\bm{\theta}_d})
\\ + \sum_{ d} \left[ \frac{N}{2}\log(2\pi) 
+ \log \det(\bm{\Sigma}_{ \bm{\theta}_d}) \right]
\end{gathered}
\label{nll}
\end{equation}
where the weights $\bm{w}_d$ are encoded in $\bm{\mu}_{\bm{\theta}_d}$, and the rest of the parameters are encoded in $\bm{\Sigma}_{\bm{\theta}_d}$.
It is worth noting that the noise $\sigma_n^2$ in displacement-based representation is the same for all policies; while in arc-based representation, Eq. (\ref{covariance_matrix}) needs to be modified:
\begin{equation}
\bm{\Sigma} = \bm{K}(\bm{x_{1:N}}, \bm{x_{1:N}}) + \sigma_{n}^2 \bm{D}
\label{modified_covariance_matrix}
\end{equation}
where matrix $\bm{D}$ is a diagonalized matrix with entries calculated according to Eq. (\ref{noises_for_v_w_phi}).
The optimal solution of Eq. (\ref{MLE}) can be obtained using optimization algorithms like L-BFGS-B. 
The Monte Carlo sampling seems computationally expensive. But note that the sample amount for each environment is small and invariant to the size of the repertoire.
Hence it is much cheaper to do than finding the basis for the prior mean function, where we need to evaluate all the policies in several environments.
Since we have used Monte Carlo sampling to represent the prior, the integrals in Eq. (\ref{integral_2}) and Eq. (\ref{integral_1}) now become averages over finite number of terms:
\begin{equation}
\begin{gathered}
p(y_i|c_i, \bm{y}_{-i}, \bm{c}_{-i}) = 
\sum_{l=1}^L p(y_i|\bm{\theta}^{(l)})
p(\bm{\theta}^{(l)}|c_i, \bm{y}_{-i}, \bm{c}_{-i})
\\
p(y_i|c_i \neq j \, \text{for any} \, n_j \neq 0) = 
\frac{1}{L} \sum_{l=1}^L p(y_i|\bm{\theta}^{(l)})
\end{gathered}
\label{MC_prior_integral}
\end{equation}
Where the posterior distribution is:
\begin{equation}
p(\bm{\theta}^{(j)}|c_i, \bm{y}_{-i}, \bm{c}_{-i}) = 
\frac{
\prod_{c_k=c_i, k \neq i} p(y_k|\bm{\theta}^{(j)})}
{\sum_{l=1}^L \prod_{c_k=c_i, k \neq i} p(y_k|\bm{\theta}^{(l)})}
\label{MC_posterior}
\end{equation}
In practice, it is much easier and numerically stable to work with logarithm likelihoods as in Eq. (\ref{nll}), as the direct product of likelihoods will very likely lead to overflow.

Although the integrals are now made tractable, this method has two fatal shortages.
First, the Monte Carlo sampling only provides limited precision of the parameters.
As each cluster grow larger, Eq. (\ref{MC_posterior}) implies that the set of parameters that is closest to the ground truth will eventually make up 100\% of the posterior weight.
Hence, the number of clusters will not exceed the amount of sampled priors. 
Second and more importantly, the prior collection is made in simulation, which ensures no extrapolation to the real-world dynamics.
To solve these two shortages, we introduce a trick called posterior convergence estimate.
As a cluster gathers more data, the posterior distribution of its parameters will become narrower and more centred closer to the ground truth (see Fig \ref{posterior_convergence}).
When the cluster size is large, we can safely assume that the posterior becomes a sharp spike.
Hence, we may treat it as a Dirac delta distribution, and Eq. (\ref{integral_1}) becomes:
\begin{equation}
\begin{gathered}
p(y_i|c_i, \bm{y}_{-i}, \bm{c}_{-i}) \approx 
\int p(y_i|\bm{\theta})
\delta (\bm{\theta} - \bm{\theta}^*)
d\bm{\theta}
\\
= p(y_i|\bm{\theta}^*)
\end{gathered}
\label{posterior_convergence_estimate}
\end{equation}
This is called posterior convergence estimate, and the center of this Dirac delta is located using MAP. 
This means we refit the parameters for large clusters (with data size above a threshold, e.g. 30) as their posterior.
This design is to leverage the fact that for small clusters, their posteriors haven't converged and spread across the parameter space.
Hence the integrals can be efficiently approximated by Eq. (\ref{MC_prior_integral}).
While for large clusters, their posterior cannot be well represented by any prior sample, so we allow them to determine the best parameters based on their data.
The only thing that such prior should satisfy is that it covers a variety of possible cases.
Thus, the use of Monte Carlo sampling for the prior acts like a guide for smaller clusters, but neither the coverage of parameters nor the number of clusters is constrained by the sampled prior. 
\begin{figure}[h]
\centering
\includegraphics[width=0.45\textwidth]{posterior_convergence.pdf}
\caption{Example of the convergence of parameter posterior in DPMM.
Data are sampled from a infinite mixture of Gaussians, where the variance is fixed at 4, and the mean is given by a Gaussian mixture as the prior.
As the cluster grows larger, the posterior of the mean becomes more narrowed and centred at the ground truth (5.5).
}
\label{posterior_convergence}
\end{figure}

Because we have no access to the analytical form of the prior to do MAP, we can use MLE instead (for large data size, MLE and MAP gives similar results).
Hence, during Gibbs sampling, the likelihood of an episode belonging to a large cluster is calculated according to Eq. (\ref{posterior_convergence_estimate}), while the likelihood of belonging to a small cluster is calculated using the first equation in Eq. (\ref{MC_prior_integral}).
As mentioned previously, the episode needs to be removed from its current cluster which will affect the posterior when calculating the likelihood of this episode staying in its current cluster.
For small cluster, we can easily recalculate the posterior using Eq. (\ref{MC_posterior}).
For large cluster, if the data size is still large enough without this episode, the parameter $\bm{\theta}^*$ needs to refitted with this episode removed.
If the data size is no longer large enough without this episode, the posterior is calculated in the same way as a small cluster.
After each resample step, we also need to refit the parameters for large clusters if any change is made.
Be aware that the subscript $i$ in Eq. (\ref{MC_prior_integral}), Eq. (\ref{MC_posterior}) and Eq. (\ref{posterior_convergence_estimate}) refers to each episode instead of each individual interaction data. 
Similarly, the $N$ and $n_j$ in Eq. (\ref{indicator_posterior_3}) correspond to the number of episodes instead of the amount of data.

In practice, refitting all the parameters in each step of Gibbs sampling could lead to performance issues as the Gibbs sampling cannot be paralleled.
It is much cheaper to just refit the weights for the prior mean function as they are most important can be obtained analytically.
Hence, for a large cluster, we use the most likely kernel in the Monte Carlo sampled prior and calculate the best fitting weights using:
\begin{equation}
\bm{w}_d^* = \left(\sum_{i=1}^n
\bm{X}_i^T \bm{\Sigma}_{i, \bm{\theta}_d}^{-1} \bm{X}_i
\right)^{-1} \sum_{i=1}^n \left( \bm{X}^T_i \bm{\Sigma}_{i, \bm{\theta}_d}^{-1} \Delta \bm{x}_{i, d} \right)
\label{refitted_weights}
\end{equation}
This is the formula we use to determine the most suitable prior for each cluster.
The summation is made over all the episodes in the cluster.
$\Delta \bm{x}_{i, d}$ is the real-world distortions of the $i^{\text{th}}$ episode.
$\bm{X}_i$ is the coordinate matrix of this episode, where each row gives the coordinates of the policy corresponding to each interaction.
The covariance matrix $\bm{\Sigma}_{i, \bm{\theta}_d}$ is generated using the most likely kernel.

To sum up, the complete algorithm for conducting DP clustering is presented in the pseudo code below:
\begin{algorithm}
\caption{DP Clustering}
\begin{algorithmic}
\STATE \textbf{procedure} DP Clustering
\STATE $\mathcal{R} \leftarrow $ MAP{\_}Elites() \COMMENT{Generate the repertoire}
\STATE $\bm{X} \leftarrow \emptyset$ \COMMENT{Initialize collected distortions}
\FOR{$\bm{i} = 1 \rightarrow G$}
\STATE $\varepsilon \leftarrow$ random{\_}simulation{\_}settings()
\STATE $\Delta \bm{x}_{1:D} \leftarrow $ evaluate{\_}repertoire($\mathcal{R}$, $\varepsilon$)
\STATE $\bm{X}$.append($\Delta \bm{x}_{1:D}$)
\ENDFOR
\STATE $\bm{X} \leftarrow$ SVD($\bm{X}$) \COMMENT{Reduce the dimensionality via SVD}
\STATE $\bm{X} \leftarrow$ PCA($\bm{X}$) \COMMENT{Convert into input space via PCA}
\STATE

\STATE $H \leftarrow \emptyset$ \COMMENT{Initialize Monte Carlo sampled priors}
\FOR{$\bm{i} = 1 \rightarrow N$}
\STATE $\varepsilon\leftarrow$ random{\_}simulation{\_}settings()
\STATE $\mathcal{D} \leftarrow \emptyset$ \COMMENT{Initialize dataset}
\FOR{$\bm{j} = 1 \rightarrow M$} 
\STATE $\Theta \leftarrow$ evaluate{\_}random{\_}policy()
\STATE $\mathcal{D}$.append($\Theta$)
\ENDFOR
\STATE $\bm{\theta} \leftarrow$ MLE($\mathcal{D}$) \COMMENT{Find the parameters with MLE}
\STATE $H$.append($\bm{\theta}$)
\ENDFOR
\STATE \COMMENT{Gibbs sampling}

\STATE $\xi^{*} \leftarrow -\infty$ \COMMENT{Initialize highest likelihood}
\STATE $\mathcal{C} \leftarrow$ initialize{\_}cluster{\_}assignment() 
\FOR{$\bm{i} = 1 \rightarrow T$} 
\STATE $\mathcal{C}$, $\xi$ $\leftarrow$ Gibbs{\_}sampling{\_}sweep($\mathcal{C}$, $H$)
\IF{$\xi > \xi^{*}$}
\STATE $\xi^{*} \leftarrow \xi$
\STATE $\mathcal{C}^{*} \leftarrow \mathcal{C}$ \COMMENT{Record the best result}
\ENDIF
\ENDFOR
\RETURN $\mathcal{C}^{*}$ 
\end{algorithmic}
\label{DP_clustering}
\end{algorithm}
In our implementation, we modelled the clustering and online-adaptation to be made in two separated systems.
The DP clustering of historical data is conducted in the archive management system, which would typically be a cloud server where expensive data processing is available.
After having found the clusters, the results are saved as an adaptation system to be used (downloaded) by individual robots.
The adaptation system is independent from the archive, and we are free to make simplifications for embedded systems.

\subsection{Learning the Dynamics Online}
In theory, the dynamics is modelled with an infinite mixture of GPs, where the GP parameters follow a finite mixture of distributions:
\begin{equation}
p(\bm{\theta}) = \sum_j \frac{n_j}{n + \alpha} p(\bm{\theta}|c=j) 
+ \frac{\alpha}{n + \alpha} H(\bm{\theta})
\label{trained_mixture_model}
\end{equation}
where the summation is made over the clusters containing at least one episode, and $p(\bm{\theta}|c=j)$ denotes the posterior of parameters for the $j^{\text{th}}$ cluster.
The last term results from the important property of DPMM that there is always a non-zero probability that we will get a new cluster (face a new situation).
Each instance of $\bm{\theta}$ sampled from $p(\bm{\theta})$ is a certain configuration of parameters of GP, and the learned distortion is integrated over the posterior of $\bm{\theta}$ conditioned on the real-world interactions collected online:
\begin{equation}
\begin{gathered}
p(x^*_d|\bm{x}_{1:N}) 
\\
= \sum_j p(c=j| \bm{x}_{1:N}) \int p(x^*_d|\bm{\theta}, \bm{x}_{1:N})p(\bm{\theta}|c=j, \bm{x}_{1:N}) d\bm{\theta} 
\\
+ p(c \neq j, \forall j|\bm{x}_{1:N}) \int p(x^*_d|\bm{\theta}, \bm{x}_{1:N}) p(\bm{\theta} | \bm{x}_{1:N}) d\bm{\theta}
\end{gathered}
\label{MGP_posterior}
\end{equation}
where the indicator probabilities are given by:
\begin{equation}
\begin{gathered}
p(c=j| \bm{x}_{1:N}) 
= \frac{n_j \cdot p(\bm{x}_{1:N}|c=j)
}{Z}
\\ 
p(c \neq j, \forall j| \bm{x}_{1:N})
= \frac{\alpha \cdot p(\bm{x}_{1:N}|c \neq j, \forall j)
}{Z}
\end{gathered}
\label{MGP_indicator}
\end{equation}
where $\bm{x}_{1:N}$ stands for the collected ($N$ number of) interaction data during deployment.
